Simple connectome inference from partial correlation statistics in calcium imaging

### 2014-06-17 article.tex
Abstract + Intro

   %\documentclass[wcp,gray]{jmlr} % test grayscale version
\documentclass[wcp]{jmlr}

%% Personnal package
\usepackage{enumerate}
\usepackage{bbold}
\usepackage{wrapfig}

 % The following packages will be automatically loaded:
 % amsmath, amssymb, natbib, graphicx, url, algorithm2e

 %\usepackage{rotating}% for sideways figures and tables
% \usepackage{longtable}% for long tables

 % The booktabs package is used by this sample document
 % (it provides \toprule, \midrule and \bottomrule).
 % Remove the next line if you don't require it.
\usepackage{booktabs}
 % The siunitx package is used by this sample document
 % to align numbers in a column by their decimal point.
 % Remove the next line if you don't require it.
\usepackage[load-configurations=version-1]{siunitx} % newer version
 %\usepackage{siunitx}
\usepackage[utf8]{inputenc}

@@

% % Define an unnumbered theorem just for this sample document:
%\theorembodyfont{\upshape}
%\theoremheaderfont{\scshape}
%\theorempostheader{:}
%\theoremsep{\newline}
%\newtheorem*{note}{Note}

 % change the arguments, as appropriate, in the following:
\jmlrvolume{1}
\jmlryear{2014}
\jmlrworkshop{\textcolor{red}{Neural Connectomics Workshop}}

% \title[Connectomics challenge]{Inferring neural networks from fluorescent
%                                calcium imaging using partial correlation and
%                                sample weighting \textcolor{red}{plus court} : Simple and robust inference of connectomes using partial correlation coefficients
% ?}

% \title[Inference of connectomes using partial correlation coefficients]{Simple and robust inference of connectomes using partial correlation coefficients}

\title{\textcolor{red}{Simple} connectome inference using partial correlation from imaging calcium signal}

% %%
% Discovery/Infering/Retrieving/... connectomes
% Robust
% Using partial correlation
% [from calcium imaging]

 % Use \Name{Author Name} to specify the name.
 % If the surname contains spaces, enclose the surname
 % in braces, e.g. \Name{John {Smith Jones}} similarly
 % if the name has a "von" part, e.g \Name{Jane {de Winter}}.
 % If the first letter in the forenames is a diacritic
 % enclose the diacritic in braces, e.g. \Name{{\'E}louise Smith}

 % Two authors with the same address
%  \author{\Name{Author Name1\nametag{\thanks{with a note}}} \Email{abc@sample.com}\and
%   \Name{Author Name2} \Email{xyz@sample.com}\\
%   \addr Address}

 % Three or more authors with the same address:
 % \author{ Antonio Sutera %\and
 %         % \Name{Arnaud Joly} \and
 %         % \Name{Vincent Fran\c{c}ois-Lavet} \and
 %         % \Name{Aaron Qiu} \and
 %         % \Name{Gilles Louppe} \and
 %         % \Name{Damien Ernst} \and
 %         % \name{Pierre Geurts}}
 %         }
\author{Antonio Sutera,
        Arnaud Joly,
        Vincent François-Lavet,
        Aaron Qiu, \\
        Gilles Louppe,
        Damien Ernst,
        Pierre Geurts}

@@

\begin{abstract}
Understanding how the brain works is a key
to treat brain disorders such as Parkinson's disease,
epilepsy. Retrieving connectomes, the neurons wiring map, will shed a new
light on the anatomical and functional connectivity of the brain. In the
context of the Connectomics challenge, we propose a simple algorithm made of
four stages, using preprocessing and partial correlation to achieve a network
inference of high quality. An optimized variation of this simple algorithm
is the winning method of the Connectomics Challenge.
We put in perspective our method
with other inference methods such as GTE (\cite{stetter2012model}) and Genie3
(\cite{huynhthu2010inferring}).

% Discussion on the metrics?
%\textcolor{red}{Discussion on the order of the partial correlation}

\end{abstract}

\begin{keywords}
Network inference - Partial correlation - Precision matrix - GENIE3
\end{keywords}

@@

% BG : brain, complex organ
The human brain is a complex biological organ, formed by 100
billions of neurons with 7000 synaptic connections on average.
Cutting edge optical devices delivers in real-time  2-D  images of
the neuronal activity of thousands of neurons through a fluorescent
calcium indicator [todo cite]. Automatic and analytical methods are required to infer
the connectomes from those time series and have to handle several
experimental issues. Some neurons of the network might not be
observed or are confounded with others. The lower device sampling rate with
respect to neuronal activity speed might not allow to distinguish neuron
activation order, especially in highly synchronous activity
[Stetter and in Stetter [20,46,47]] of a large part of the network. Moreover
the fluorescent concentration is highly reactive to a neuron activation or a
spike, but have a slow decay.

% Introduce notation + goal in mathematical score
The connectomes, the neural network, is represented by a (un)directed graph
$G = (V, E)$ comprising
a set of nodes $V$ and a set of edges $E \subseteq \left\{(i, j) \in V \times
V\right\}$, (un)ordered pairs from $V$.
Each node $i \in V$ represents a neuron, and each edge $(i, j) \in E$
represents a direct link and interaction between neuron $i$ and $j$.
In the directed graph, edges also express causality: the
state of a neuron $i$ might be due to the state of one or several of its
parents $\left\{j : (j, i) \in E, j \in V\right\}$.
The network inference task is set as follows :
\textit{Given the observations $(x^t \in \mathcal{R}^{|V|})_{t=1}^T $
along $T$ periods of time of the $|V|$ neurons, the goal is to infer the
adjacency matrix $A : a_{i,j} = \mathbb{1}((i, j) \in E)$ of the neural network
of graph $G$, where $\mathbb{1}$ is the indicator function.}

% Solution
In this paper, we describe the winning method of the Connectomics challenge
\footnote{\url{http://connectomics.chalearn.org}}, a simple and theoretically
grounded approach, based on partial correlation\footnote{Source code is available
at \url{https://github.com/asutera/kaggle-connectomics}.} to infer the neural
network. The rest of the paper is structured as follows: Section \ref{sec:filter}
describes the pre-processing applied on fluorescent calcium timeseries;
Section \ref{sec:inference} presents the proposed approach and
their theoretical properties; Section \ref{sec:results} provide
empirical analysis and comparison with other network inference methods,
whereas Section \ref{sec:conclusion} discusses our work and provides further
research directions.

@@

\section{Network inference through weighted average partial correlation
         coefficients}
\label{sec:inference}

@@

\section{Conclusion} \label{sec:conclusion}

@@

\begin{scriptsize}

\paragraph{Acknowledgements.} A. Joly and G. Louppe are research fellows of
the FNRS, Belgium.  A. Sutera is  is recipient of
a F.R.I.A. fellowship of F.R.S.-FNRS, Belgium.
This work is supported by PASCAL2 and the IUAP DYSCO, initiated by the
Belgian State, Science Policy Office.

\end{scriptsize}

>>>>

\documentclass[wcp]{jmlr}

\usepackage{enumerate}
\usepackage{bbold}
\usepackage{booktabs}
\usepackage{natbib}
\usepackage[utf8]{inputenc}

@@

 % change the arguments, as appropriate, in the following:
%\jmlrvolume{1}
%\jmlryear{2014}
%\jmlrworkshop{\textcolor{red}{Neural Connectomics Workshop}}
\jmlrproceedings{}{}

\title{Simple connectome inference from partial correlation statistics in calcium imaging}

\author{Antonio Sutera,
        Arnaud Joly,
        Vincent François-Lavet,
        Aaron Qiu, \\
        Gilles Louppe,
        Damien Ernst,
        Pierre Geurts}

@@

\begin{abstract}
In this work, we propose a simple and mathematically motivated method for the
problem of connectome inference in calcium imaging data. The proposed algorithm
consists in i) processing the raw signals to detect neural peak activities and ii)
inferring the degree of association between neurons from partial correlation
statistics.  This paper summarizes the methodology that led us to win the
Connectomics Challenge, proposes a simplified version of our method and finally
discusses our results with respect to other inference methods.
\end{abstract}

\begin{keywords}
Connectomics - Network inference - Partial correlation
\end{keywords}

@@

The human brain is a complex biological organ made of 100 billions of neurons,
each connected to 7000 other neurons on average. Unfortunately, direct
observations of the connectome, the wiring diagram of the brain, is not yet
technically feasible. Without being perfect, calcium imaging currently allows
the realtime and simultaneous observation of neuron activities from thousands
of neurons, producing individual time series representing their fluoresence
intensity. In this context,  the connectome inference problem consists in
retrieving the synaptic connections between neurons on the basis of the
fluoresence time series. In particular, this problem is often made difficult
because of experimental issues, including masking effects making some of the
neurons not to be observed or confounded with others, a low sampling rate of
the optical device with respect to the neural activity speed or the slow decay
of fluoresence.

Formally, the connectome can be represented as a directed graph $G=(V,E)$,
where $V$ is a set of $p$ nodes, representing neurons, and $E \subseteq
\left\{(i, j) \in V \times V\right\}$ is a set of edges, representing direct
synaptic connections between neurons. Causal interactions are expressed by the
direction of edges: $(i, j) \in E$ indicates that the state of neuron $j$ might
be caused by the activity of neuron $i$. In those terms,  the connectome
inference problem is formally stated as follows:  \textit{Given the sampled
observations $\{ x^t_i \in \mathbb{R}^{p} | i \in V, t = 1, \dots T \}$ of $p$
neurons for $T$ time intervals, the goal is to infer the set $E$ of connections in $G$.}

In this paper, we describe the winning method\footnote{Code available at
\url{https://github.com/asutera/kaggle-connectomics}} of the Connectomics
Challenge\footnote{\url{http://connectomics.chalearn.org}}, a simple and
theoretically grounded approach based on signal processing techniques and
partial correlation statistics to infer the neural network. The rest of the
paper is structured as follows: Section \ref{sec:filter} describes the signal
processing methods applied on fluorescent calcium time series; Section
\ref{sec:inference} then presents the proposed approach and its theoretical
properties; Section \ref{sec:results} provides an empirical analysis and
comparison with other network inference methods, while Section
\ref{sec:conclusion} finally discusses our work and provides further research
directions.

@@

\section{Connectome inference from partial correlation statistics}
\label{sec:inference}

@@

\section{Conclusions} \label{sec:conclusion}

@@

\section*{Acknowledgements}

A. Joly and G. Louppe are research fellows of
the FNRS, Belgium.  A. Sutera is  is recipient of
a F.R.I.A. fellowship of F.R.S.-FNRS, Belgium.
This work is supported by PASCAL2 and the IUAP DYSCO, initiated by the
Belgian State, Science Policy Office.
